TacticToe: Learning the Game of Tactics

### 2018-01-22 tactictoe_jv.tex
record: simplification

@@

\subsection{Minimal \sml Code Trees}
\holfour proofs are organized in theory files.
Each theory file contains definitions and statements of theorems together with
their proofs. Those theories and the contained proofs
are written
in the functional programming language \sml. To automate
proof search strategies a user can implement programming language functions
that correspond to more involved proofs.
Functions that correspond to backward proofs, called tactics, are combined with each other with the help of 
tacticals (functions that takes
tactics as arguments). Such a combination of tactics will be called a \emph{proof script} when it
proves a theorem stated in a theory.

@@

Using the standard types $string$ and $list$ and their type constructors,
we define a minimal code tree by the following AST:
\begin{align*}
M &=_{def} string \\
I &=_{def} string\ \vert\ Module(M,I)\\
D &=_{def} Open(M)\ \vert\  Val(I,E) \\
E &=_{def} Code(I\ list)\ \vert\   Let (D,E)
\end{align*}

In this AST, the constructors $Let$, $Val$ and $Open$ correspond to
the reserved \sml tokens \texttt{let}, \texttt{val} and \texttt{open} 
respectively. $Module(M,I)$ refers to the \sml token \texttt{M.I} where 
\texttt{M} is a module (also called structure in \sml) and \texttt{I} is an 
identifier. $D$ is a declaration, $E$ an expression. Inside a 
declaration $val(I,E)$, $E$ is a definition for $I$.
To construct such an AST for a theory file, we rely on a custom parser. It transform a source file into a 
minimal code tree taking into account the scope of each constructor.
An additional reason to refrain from the use of the interpreter in our case, is that
the \polyml compiler needs to execute the code preceding an expression 
before parsing it, which is significantly slower than a custom parser.

\subsection{Globalization}
To transfer \sml values as code across theories in a conservative manner, we 
need to 
globalize \sml identifiers. To this end, we replace each identifier by some 
\sml code containing all the information to obtain the value of this identifier 
without knowing the context. The result is a standalone \sml code executable in 
any \holfour theories which will be interpreted in the same way in the absence 
of side effects. Two techniques are helpful to produce such code. The first one 
is to prefix global identifiers by the module (and sub-modules) in which they 
were declared. And the second one is to replace local identifiers 
by their definitions. Prefixing and 
unfolding are applied recursively in 
the definitions.

For a more formal explanation, we define a target AST build upon the original 
one, where each identifer $I$ is replace by its globalized expression $E'$:

\begin{align*}
M' &=_{def} string \\
I' &=_{def} string\ \vert\ Module(M',I')\\
D' & =_{def} Open(M')\ \vert\  Val (I',E') \\
E' &=_{def} Code(E'\ list)\ \vert\ Let (D',E') 
\end{align*}

Then, we express two mutually recursive functions $f_d$ and 
$f_e$ that globalize the original representation. The 
function $f_d$ modifies declaration and $f_e$ modifies expressions. Since a 
\sml program is a list of declarations, we will be able transform it by 
iteratively applying $f_d$. The stack 
$\mathbb{S}$ keep track of each previous declaration: the identifier that is 
declared and their globalized definition. To save a declaration to 
the stack, we use the $push$ 
function. Opening a module has the effect of pushing all identifiers with 
their associated prefixed version to the stack. The scope of an identifier is 
automatically handled by construction of the AST. To find the globalized 
version of an identifier $i$, we use the $lookup$ function on a stack $s$, 
which returns the globalization associated with $i$ appearing highest in $s$. 
We then build $f_e$ and $f_d$ in the following way:
\vspace{5mm}

\begin{math}
\begin{array}{lcl}
f_d:& \mathbb{S} \times D &\rightarrow\ \ \mathbb{S} \times D'\\
& (s, Open(m)) &\mapsto\ \ (push(m,s),\ Open(m)) \\
& (s, Val(i,e)) &\mapsto\ \ (push ((i,\hat{e}),s),\ Val(i, \hat{e})) 
\mbox{ where } \hat{e} = \pi_2^2 (f_e (s,e))
\end{array}
\end{math}

\begin{math}
\begin{array}{lcl}
f_e:& \mathbb{S} \times E &\rightarrow\ \ \mathbb{S} \times E'\\
& (s, List(l))  &\mapsto\ \ (s,\ lookup(s,l))\\
& (s, Let (d,e)) &\mapsto\ \ (\hat{s}, f_e (\hat{s}, e))
\mbox{ where } \hat{s} = \pi_1^2 (f_d (s,d))
\end{array}
\end{math}

@@

Thanks to the simplicity of the minimal AST, we are able to present our 
recursive unfolding algorithm in a concise and formal manner. But for practical 
purpose, the minimal AST is not sufficient. So, we extend the scope of our 
algorithm by enumerating common \sml constructions and explain 
independently how we globalize them. We first start by mentioning additional 
details about module and value declarations.

@@

\paragraph{Values}
The standard ML language allows to declare multiple values in a single $let$ 
construction. In effect, this list of declarations is equivalent to nested 
declarations.

Because we do not want to unfold proofs into other proofs, we 
make an exception 
for values of type theorem and do not replace them by their definition.
Instead, we fetch theorems from the 
database by looking for its formula in the database of theorems.
Because this database is reconstructed from files saved on disk, a call to 
the function \texttt{DB.fetch} works across theories.

\paragraph{Functions}
To globalize a function identifier, we include 
its declaration in the globalized expression. 
Given a declaration \texttt{fun f v = e}, the globalized 
expression for the identifier \texttt{f} is \texttt{let fun f v = e' in f 
end}, where \texttt{e'} is the globalized version of \texttt{e}. The function 
\texttt{f} and its argument \texttt{v} are binders. So, we prevent our 
algorithm from trying to globalize them during the construction of \texttt{e'}.
This approach is slightly more verbose than using an 
anonymous function construction. But it is prettier as 
we regroup declarations at the top of the proof expression before returning it
to the user (see Section~\ref{sec:proofdisplay}).

\paragraph{Pattern matching}
A pattern is an \sml expression that 
contains \sml 
constructors (or exceptions in the case of \texttt{handle}) and binding 
variables. Our algorithm can recognize cases where pattern matching appears and 
globalize the expression appropriately.  Let \texttt{p} be a pattern 
and 
\texttt{e} be abd expression, the following constructions can 
occur: \texttt{val p = e}, \texttt{case \_ of p => e}, \texttt{fun f p = e}, 
\texttt{fn p => e}, \texttt{handle p => e}. A binding identifiers in \texttt{p}
is a token that is not a constructor (or exception). To produce the globalized 
version of each construction, we globalize constructors (or exception) in the 
pattern \texttt{p} and we protect binding variables in \texttt{e} to prevent 
their modification. When a value identifier \texttt{x} is declared in a pattern 
\texttt{p}, the globalized expression \texttt{e'} for \texttt{x} becomes 
\texttt{let val p = e' in x end}.

\paragraph{Infixity}
In \sml, it is possible to declare infixity rules for a string $s$ that 
is not prefixed by a module. The infixity applies only when an identifier 
exactly match $s$. Therefore, an identifier \texttt{x} could have been treated 
as infix in an original expression \texttt{E} but when 
we substitute it locally by its 
globalized version \texttt{e'}, \texttt{e'} is now considered with prefix 
infixity in \texttt{E} which makes \texttt{E} an invalid expression. Therefore, 
because we 
want to avoid global changes to \texttt{E}, we encapsulate \texttt{e'} by two 
newly 
defined 
infix guards to reproduce 
the original behaviour.
In total there are 20 possible infix settings \texttt{x} can have in \sml.
Those infix guards are defined to have the 
same 
precedence level $\texttt{P}\in[0,9]$ and same associativity 
$\texttt{A}\in[l,r]$ (right or left) as \texttt{x}.
Definitions for the infix guards depend on their associativity:

\begin{definition}(Code of infix guards)
\begin{lstlisting}
fun a infixlX_open f = fn x => f (a,x)  
fun g infixlX_close b = g b

fun f infixrX_close b = fn x => f (x,b)
fun a infixrX_open g = g a
\end{lstlisting}
\end{definition}

\begin{proof}(Does anyone want a proof that this functions are correctly 
reproducing the original function?)
\end{proof}

@@

\subsection{Start of the search}
The input of the algorithm is a goal $g$ that we want to prove.
Therefore the search tree starts with only one node containing the singleton 
$g$ and an ordered list of predicted tactics for the goal $g$.

@@

\subsection{Node extension}
Once a node is selected, we applied the best not-yet-applied tactics to
the open goal this node. If the tactic is productive then the we create a new 
node and calculate and save its prior evaluation. Otherwise, we create a 
non-accessible node that holds a value of 0 to notice the value function that 
node selection was a failure. This dummy node is only there to update the 
result of the Current evaluation of other nodes and is not considered in the 
calculation of exploration term and cannot be selected.

>>>>

\subsection{The problem}

@@

%\subsection{Minimal \sml Code Trees}
%
%
%Using the standard types $string$ and $list$ and their type constructors,
%we define a minimal code tree by the following AST:
%\begin{align*}
%M &=_{def} string \\
%I &=_{def} string\ \vert\ Module(M,I)\\
%D &=_{def} Open(M)\ \vert\  Val(I,E) \\
%E &=_{def} Code(I\ list)\ \vert\   Let (D,E)
%\end{align*}
%
%In this AST, the constructors $Let$, $Val$ and $Open$ correspond to
%the reserved \sml tokens \texttt{let}, \texttt{val} and \texttt{open} 
%respectively. $Module(M,I)$ refers to the \sml token \texttt{M.I} where 
%\texttt{M} is a module (also called structure in \sml) and \texttt{I} is an 
%identifier. $D$ is a declaration, $E$ an expression. Inside a 
%declaration $val(I,E)$, $E$ is a definition for $I$.
%To construct such an AST for a theory file, we rely on a custom parser. It 
%transform a source file into a 
%minimal code tree taking into account the scope of each constructor.
%An additional reason to refrain from the use of the interpreter in our case, 
%is that
%the \polyml compiler needs to execute the code preceding an expression 
%before parsing it, which is significantly slower than a custom parser.

\subsection{Globalization}

\holfour proofs are organized in theory files.
Each theory file contains definitions and statements of theorems together with
their proofs. Those theories and the contained proofs
are written
in the functional programming language \sml. To automate
proof search strategies a user can implement programming language functions
that correspond to more involved proofs.
Functions that correspond to backward proofs, called tactics, are combined with 
each other with the help of 
tacticals (functions that takes
tactics as arguments). Such a combination of tactics will be called a 
\emph{proof script} when it
proves a theorem stated in a theory.

@@

%\subsubsection {Unfolding locally declared values}
To transfer \sml values as code across theories in a conservative manner, we 
need to 
globalize \sml identifiers. To this end, we replace each identifier by some 
\sml code containing all the information to obtain the value of this identifier 
without knowing the context. The result is a standalone \sml code executable in 
any \holfour theories which will be interpreted in the same way in the absence 
of side effects. Two techniques are helpful to produce such code. The first one 
is to prefix global identifiers by the module (and sub-modules) in which they 
were declared. And the second one is to replace local identifiers 
by their definitions. Prefixing and 
unfolding are applied recursively in 
the definitions.

%For a more formal explanation, we define a target AST build upon the original 
%one, where each identifer $I$ is replace by its globalized expression $E'$:
%
%\begin{align*}
%M' &=_{def} string \\
%I' &=_{def} string\ \vert\ Module(M',I')\\
%D' & =_{def} Open(M')\ \vert\  Val (I',E') \\
%E' &=_{def} Code(E'\ list)\ \vert\ Let (D',E') 
%\end{align*}
%
%Then, we express two mutually recursive functions $f_d$ and 
%$f_e$ that globalize the original representation. The 
%function $f_d$ modifies declaration and $f_e$ modifies expressions. Since a 
%\sml program is a list of declarations, we will be able transform it by 
%iteratively applying $f_d$. The stack 
%$\mathbb{S}$ keep track of each previous declaration: the identifier that is 
%declared and their globalized definition. To save a declaration to 
%the stack, we use the $push$ 
%function. Opening a module has the effect of pushing all identifiers with 
%their associated prefixed version to the stack. The scope of an identifier is 
%automatically handled by construction of the AST. To find the globalized 
%version of an identifier $i$, we use the $lookup$ function on a stack $s$, 
%which returns the globalization associated with $i$ appearing highest in $s$. 
%We then build $f_e$ and $f_d$ in the following way:
%\vspace{5mm}
%
%\begin{math}
%\begin{array}{lcl}
%f_d:& \mathbb{S} \times D &\rightarrow\ \ \mathbb{S} \times D'\\
%& (s, Open(m)) &\mapsto\ \ (push(m,s),\ Open(m)) \\
%& (s, Val(i,e)) &\mapsto\ \ (push ((i,\hat{e}),s),\ Val(i, \hat{e})) 
%\mbox{ where } \hat{e} = \pi_2^2 (f_e (s,e))
%\end{array}
%\end{math}
%
%\vspace{5mm}
%
%\begin{math}
%\begin{array}{lcl}
%f_e:& \mathbb{S} \times E &\rightarrow\ \ \mathbb{S} \times E'\\
%& (s, List(l))  &\mapsto\ \ (s,\ lookup(s,l))\\
%& (s, Let (d,e)) &\mapsto\ \ (\hat{s}, f_e (\hat{s}, e))
%\mbox{ where } \hat{s} = \pi_1^2 (f_d (s,d))
%\end{array}
%\end{math}
%
%\vspace{5mm}

@@

%Thanks to the simplicity of the minimal AST, we are able to present our 
%recursive unfolding algorithm in a concise and formal manner. But for 
%practical 
%purpose, the minimal AST is not sufficient.

We present the tactical proof recording mechanism in details here because it 
can help 
another researcher to produce a tactical export in other interactive theorem 
provers.
We now enumerate common \sml constructions and explain 
independently how we globalize them. We use a stack to keep track of previous 
identifier declaration and their globalization.

%We first start by mentioning additional 
%details about module and value declarations.

\subsection{Storing Declaration}

\paragraph{Reserved tokens}

\paragraph{Module}

\paragraph{Global identifier}

\paragraph{Local identifier}

To globalize a identifier, we include 
the code of its declaration in the globalized expression. 
Given a declaration \texttt{fun f v = e} or \texttt{val f = e}, the globalized 
expression for the identifier \texttt{f} is \texttt{let fun f v = e' in f 
end} respectively \texttt{let val f = e' in f end}
The expression \texttt{e'} is the globalized version of \texttt{e}. That is to 
say, it is the expression \texttt{e} where each of its tokens has been replaced 
its globalized expressions as constructed in Section~\ref{}.

In the case of value, we can simplify the globalized expression further and 
replace it by \texttt{e'}.
In the case of a function, \texttt{f} and its argument \texttt{v} are binders. 
So, we prevent their globalization during the construction of \texttt{e'}.
%This approach is slightly more verbose than using an 
%anonymous function construction. But it is prettier as 
%we regroup declarations at the top of the proof expression before returning it
%to the user (see Section~\ref{sec:proofdisplay}).

\paragraph{Bound variables}
A pattern is an \sml expression that 
contains \sml 
constructors (or exceptions in the case of \texttt{handle}) and binding 
variables. Our algorithm can recognize cases where pattern matching appears and 
globalize the expression appropriately.  Let \texttt{p} be a pattern 
and 
\texttt{e} be abd expression, the following constructions can 
occur: \texttt{val p = e}, \texttt{case \_ of p => e}, \texttt{fun f p = e}, 
\texttt{fn p => e}, \texttt{handle p => e}. A binding identifiers in \texttt{p}
is a token that is not a constructor (or exception). To produce the globalized 
version of each construction, we globalize constructors (or exception) in the 
pattern \texttt{p} and we protect binding variables in \texttt{e} to prevent 
their modification. When a value identifier \texttt{x} is declared in a pattern 
\texttt{p}, the globalized expression \texttt{e'} for \texttt{x} becomes 
\texttt{let val p = e' in x 
end}.

\paragraph{Infixity}                                                            
In \sml, it is possible to declare infixity rules for a string $s$ that 
is not prefixed by a module. The infixity applies only when an identifier 
exactly match $s$.

\subsection{Globalizing Expression}

\paragraph{Reserved tokens}

@@

\paragraph{Global identifier}
Global values and functions are prefixed by the modules and submodules where 
they were declared. This information obtained by looking up the first 
identifier that corresponds to the values at the top of the stack.

\paragraph{Local identifier}
Each local identifier is replaced by its globalized expression.

We do not unfold local and global values of type theorem. Indeed, we do not 
want to 
unfold proofs into other proofs. Instead, we fetch theorems by looking for its 
formula in the global database of 
theorems. If it does not exists, we consider the tactic to be local and not 
transferable to other theories.

\paragraph{Bound variables}
binding variables are not globalized

\paragraph{Infixity}
 Therefore, an identifier \texttt{x} could have been treated 
as infix in an original expression \texttt{E} but when 
we substitute it locally by its 
globalized version \texttt{e'}, \texttt{e'} is now considered with prefix 
infixity in \texttt{E} which makes \texttt{E} an invalid expression. Therefore, 
because we 
want to avoid global changes to \texttt{E}, we encapsulate \texttt{e'} by two 
newly 
defined 
infix guards to reproduce 
the original behaviour.
In total there are 20 possible infix settings \texttt{x} can have in \sml.
Those infix guards are defined to have the 
same 
precedence level $\texttt{P}\in[0,9]$ and same associativity 
$\texttt{A}\in[l,r]$ (right or left) as \texttt{x}.
Definitions for the infix guards depend on their associativity:

\begin{definition}(Code of infix guards)
\begin{lstlisting}
fun a infixlX_open f = fn x ==> f (a,x)      fun g infixlX_close b = g b
fun a infixrX_open g = g a      fun f infixrX_close b = fn x ==> f (x,b) 
\end{lstlisting}
\end{definition}

@@

\subsection{Start of the search}
The input of the algorithm is a goal $g$ that we want to prove.
Therefore the search tree starts with only one node containing the singleton 
$g$ and an ordered list of predicted tactics for the goal $g$.
Then, the MCTS algorithm alternates node selection steps and node extension 
steps which grow the search tree.

@@

\subsection{Node extension}
Once a node is selected, we applied the best not-yet-applied tactics according 
to the prediction order to the open goal this node. If the tactic is productive 
then the we create a new 
node and calculate and save its prior evaluation. Otherwise, we create a 
non-accessible node that holds a value of 0 to notice the value function that 
node selection was a failure. This dummy node is only there to update the 
result of the Current evaluation of other nodes and is not considered in the 
calculation of exploration term and cannot be selected.
